Naturalness and Long-lived Beyond the Standard Model Particles

### 2017-07-21 tex/draft.tex
Light edit of whole thing

    (*kinda funny wording throughout. Maybe it would be better to trade some words for math?)
    All SM particles interact with virtual particles through loop interactions. These interactions lead to corrections to SM parameters such as particle masses. It is common to write a given parameter as a linear combination of its 'bare' (uncorrected) value and its corrections. As the only fundamental scalar in SM, the Higgs is uniquely affected by these corrections. The next section explores the concept of naturalness, a major motivating principle for beyond the standard model (BSM) physics that is directly tied to the relationship between the Higgs bare mass and the terms that correct it.

\section{Naturalness and the Higgs mass}
    The naturalness criterion takes several forms. Here are a few (*cite these): (1) all parameters in a model should be of order 1, (2) Loop-level contributions should not be larger than tree-level contributions, (3) any small parameter should be protected by a custodial symmetry, and (4) fine-tuning is bad (*probably a better way to say 4). Although these statements are all trying to get at the same fundamental notion, I find (3) and (4) to be most compelling.

@@

    (*talk about anthropic principle or other arguments against?)
    (*I got tired of reading this section and haven't edited this paragraph at all yet)
    There are a few proposed solutions to the SM naturalness problem. Supersymmetry (SUSY) is the main one, and I'll spend the great majority of my time on it. Extra dimensions and composite Higgs models are the other two that pop up fairly often, but won't spend much time on them outside the next two sentences. Extra dimensions solves the hierarchy problem either by lowering the cutoff scale (as in large extra dimensions (ADD)) or introducing a warping factor that reduces the level of unnaturalness (warped extra dimensions (RS)). Composite Higgs models propose that the Higgs stops behaving as a fundamental scalar at some energy level, which removes the quadratic dependence. (*Hidden valley, dark sector, little Higgs?)

\section{SUSY and LLPs}
    (*is there a way to move the naming stuff elsewhere or shorten it?)
    SUSY introduces a new symmetry in which every SM field fits into a larger multiplet with an inherent symmetry between fermions and bosons. In it's simplest form, SUSY results in one new boson for every SM fermion and one new fermion for every SM boson (*and an extra Higgs field - can I gloss over that?). The increase in particle multiplicity calls for a new naming convention: the spin-0 superpartners to the SM fermions are referred to as sfermions (e.g. stop or smuon), and the spin-1/2 superpartners to the SM bosons are given the name of their SM counterpart with "ino" tacked on to the end (e.g. Higgsino or wino). Together the superpartners to the SM particles are known as sparticles. Because fermion loops come with a relative minus sign (*there must be a better way to say that), every bosonic correction to the Higgs mass is cancelled by a fermionic correction and vice versa. If supersymmetry were exact, the bare Higgs mass would be uncorrected to all orders in perturbation theory.

    Experimental evidence shows that SUSY must be broken, however, because no superpartners have been found, and exact SUSY requires each superpartner to have the same mass as its standard model counterpart. The mechanism and scale of SUSY breaking is an open question with many competing answers that achieve different theoretical goals and lead to different TeV-scale phenemena. A handful of models with particular approaches to SUSY breaking that are particularly prominant or particularly relevant to naturalness or long-lived particles (LLPs) are gauge-mediated susy breaking (GMSB), anomaly-mediated SUSY breaking (AMSB), split SUSY, and R-parity violating SUSY (RPV). The details of these models will be explored as needed throughout this paper.

    The SUSY variants most relevant to this discussion are those that are motivated by naturalness and (*and/or?) produce a signal that could be missed by common SUSY search techniques. Specifically, we are interested in SUSY models that predict LLPs. That is, particles whose lifetimes are such that they decay at least a detectable distance from the nominal interaction point, which includes everything from particles that decay at (*give ballparky number (~100um?)) to particles that propagate through the entire detector.

    In both the SM and in BSM theories, LLPs result from one of the following: (1) a symmetry that guarantees stability, (2) weak couplings to decay products, (3) highly-virtual intermediate states, or (4) limited phase space in which to decay. In the SM, the electron and proton are the lightest particles carrying electric charge and baryon number, respectively, which leads to their absolute stability through charge and baryon number conservation. Neutrons and muons, on the other hand, are long-lived beause the proton and electron masses leave them little phase space in which to decay (*also mention that they're both weak decays?). (*Mention neutrino, pion, kaon, whatever other bound state I might be forgetting?)

    In SUSY models, LLPs arise for the same basic reasons, and many different SUSY models meet different LLP criteria in different ways. The most common (*based on my assumptions) SUSY LLP occurs in models with conserved R-partity, which is a proposed symmetry that protects many models from predicting proton decay. In models with R-parity, the SM particles have R-parity +1 (*double check this explanation) and their superpartners have R-parity -1. Conserving overall R-parity (*specifically the product; does overall mean that?) at each vertex has two immediate phenomenalogical consequences: (1) sparticles can only be produced in pairs and (2) the lightest supersymmetric particle (LSP) is an absolutely stable LLP. The second phenomenon is the main motivation behind using missing energy as the typical SUSY seach signature.

    In models with R-parity violation (RPV), LLPs can be produced when the RPV terms have small coupling constants, which leads to long-lived, but not perfectly stable, LSPs (*any way non-LSPs could be long-lived for this reason too?). Luckily, RPV models can still avoid proton decay by only allowing terms that violate either baryon number or lepton number but not both. A similar situation arises in GMSB models where the gravitino is the LSP. The coupling between the next-to-LSP (NLSP) and gravitino is inversely proportional to the scale of SUSY breaking, which supresses the NSLP decay rate and can lead to LLPs(*do I need to explain GMSB here?). (*cite dimopoulos low energy)

    LLPs from highly-virtual intermediate particles can occur in models such as split (or mini-split) SUSY, where the scalar sparticles are significantly more massive than the gauginos. In these models, the gluino becomes long-lived when its decay to a quark and neutralino is mediated by a highly-virtual squark. (*cite kilian) (*mention that split throws out naturalness motivation?)

    SUSY models with degenerate sparticle spectra can have increased sparticle lifetimes from the reduced phase space for decays. In some AMSB models (*be more specific), for example, the zino and wino are the LSP and NLSP, respectively, and are nearly degenerate in mass. The wino's lifetime, then, is increased by the lack of phase space in which to decay. (*cite R\&S out of this world)
    (*merge last 2 paragraphs under 'mass spectra lead to LLP' concept?)
    (*never said anything about hidden valley or dark sector stuff. Never said the word R-hadron. Never said stealth or displaced SUSY. Should I?)

@@

\subsection{Tracker}
        (*cite CMS TDR, jinst 12 c02033)
        (*maybe I should trade some number listing for talking about how silicon detectors work)
        (*what about readout, cooling, etc?)
        In the region closest to the interaction point, CMS uses a fast, high-granularity silicon tracker to reconstruct tracks and pinpoint (*too flowery?) primary and secondary vertices. The tracker is subdivided into two regions: at r < 20 cm, the high particle flux (*specify) requires smaller pixels to keep occupancy near 0.1\%. Moving farther from the IP decreases particle flux, which allows the use of larger silicon strips.

@@

        Just outside the pixel tracker, the silicon strip tracker (SST) occupies the 20 < r < 116 cm and |z| < 282 cm region of CMS. Composed of 9.3 million silicon strips that vary from \_ to \_ in size (*look up) depending on particle flux, the SST covers an active area of roughly 196 m\^2. At each trigger, the strip output voltage is transferred via optical link to be read out by off-detector front-end electronics (*right language?).

\subsection{Electromagnetic Calorimeter}
        (*cite CMS TDR)
        (*talk about readout?)
        (*More in the way of concept/how stuff works?)
        Moving outward, the electromagnetic calorimeter (ECAL) follows the inner tracker. As a homogeneous scintillation calorimeter, ECAL uses 61200 lead tunstate crystals in the barrel and 7324 in each endcap to reconstruct the energy deposited during electromagnetic showers. Lead tungstate crystals allow for a fast (80\% of light within 25 ns), compact (radiation length = 0.89 cm), fine-grained (Moliere length = 2.2 cm), and radiation hard (up to 10 Mrad(*learn context for this number)) calorimeter. The main drawback is the relatively low light yield (30 photon / MeV), which necesitates photodetectors with intrinsic gain that work in magnetic fields (*say B-field part better).

        The barrel section (EB) extends radially from 129 to 177 (*look up) cm and covers a pseudorapidity range of 0 to 1.479. The crystals are tapered to approximately project back to the IP but not so perfectly that particle trajectories would be likely to align with cracks. Each crystal is approximately one Moliere radius wide and 25.8 radiation lengths deep.

@@

        Each endcap consists spans a pseudorapidity range of 1.3 to 3.0 with 14 towers in eta and 5 to 10 degree phi segmentation. Also, a steel and quartz fiber forward calorimeter (HF) sits 11.2 m from the interaction pipe in a pseudorapidty range of 3.0 to 5.0. In HF, particles produce Cherenkov light when traversing the quartz fibers that run parallel to the beamline. (*say why HF is useful or mention crazy flux?)

@@

\subsection{Trigger}
        (*cite CMS TDR, jinst 12 c03021)
        (*Is entire detector read out at L1?)
        (*Mention fpgas in L1 or more details about HLT?)
        (*Something about particle flow?)
        (*Sloppy, but I'm getting tired)
        The trigger reduces the data writing rate from the 40 MHz collision rate to less than 1 kHz so that events can be written to tape. The rate reduction happens in two stages: Level-1 (L1) and High-Level Trigger (HLT). The HLT uses input from ECAL, HCAL, and MS that represents physics objects (e/gamma, tau, jets, ET, and mu) and uses custom electronics to reduce the rate to ~100 kHz in 3.8 us. The HLT then uses a dedicated processor farm to reduce the rate to the desired < 1 kHz.

\section{LHC Searches for Long-lived Particles}
\subsection{Overview}
        LLPs leave distinct signatures that differentiate LLP signals from SM backgrounds but also add complexity to LLP analyses. CMS and ATLAS were generally designed to detect and reconstruct stable particles from prompt decays, so LLP analyses often need to veer off the traditional analysis path in terms of object reconstruction (*really reconstruction? Maybe object definition?), background estimation, and which data tier to use (*can I say this more generally?). This paper will focus on four complementary LLP signatures: displaced vertices, disappearing tracks, slow-moving particles, and particles that stop completely before decaying.

        Displaced vertices and disappearing tracks both look for particles that are unstable on detector timescales (*quantify?). Displaced vertices searches are sensitive to LLPs that decay to visible (*detectable? some other word?) particles in the inner volume (*specify) of the detector and leave tracks that can be traced back to a vertex that is sufficiently displaced from the nominal interaction point (*wordy). Disappearing tracks searches are sensitive to LLPs that decay to invisible (*neutral? some other word?) particles along with particles that are too soft to be reconstructed. Together, these two types of analyses cover something something parameter space of LLPs that decay on detector timescales.

        Slow and stopped particle searches are both sensitive to LLPs that are detector-stable but far more massive than any stable SM particle. Searches for low-velocity particles use ionization energy loss per unit distance (dE/dX) and time-of-flight (TOF) measurements to look for particles that are moving more slowly than the expected SM particles (*redundant), which normally (*quantify) have beta > 0.9. Stopped particle searches take this idea one step further and look for particles that actually come to rest inside the detector before decaying, a phenomenon that could be possible with long lifetimes and small initial kinetic energy. These analyses look for calorimeter energy deposits or hits in the muon system (*too specific for overview?) that are out of time with proton bunch crossings that could be due to the decay of stopped particles.

@@

\subsection{Displaced Vertices}
            (*cite CONF-2017-026)
            LLPs that decay before reaching the inner tracker but still live longer than the vast majority of SM particles can be identified by their displaced vertex (DV). That is, a reconstructed vertex far from the nominal interaction point. ATLAS performs a search for such particles by looking for at least one DV along with large MET (*has MET been defined yet?) and large track multiplicity. The search is sensitive to long-lived gluinos with lifetimes in the 20 ps to 10 ns range.

@@

            Jets and MET are reconstructed as in standard analyses, but special care must be taken when looking for displaced tracks. The standard ATLAS tracking algorithm constrains the transverse and longitudinal impact parameters of potential tracks, so the authors employ an additional algorithm, referred to as large-radius tracking (LRT). LRT is run after the standard tracking algorithm and uses only hits that were not used in standard track reconstruction. By relaxing d0, z0, and number of hits requirements, LRT is able to reconstruct tracks that point back to DVs. (*should I talk about reconstruction efficiencies?)

@@

            Both ATLAS and CMS looked for disappearing tracks during Run I, but the tracker improvements installed in ATLAS before 2015 and CMS before 2017 should increase sensitivity to decays at shorter radii (*I haven't mentioned previous iterations when talking about other analyses). The ALTAS search presented here is sensitie to LLP lifetimes of approximated 10 ps to 10 ns.

            The authors consider two signal processes: (1) electroweak gaugino production resulting in a disappearing track, an ISR jet, and MET and (2) gluino pair production resulting in a disappearing track, four jets, and MET. Process (1) could be the only gaugino production mode at the LHC if the gluino mass to kinematically unaccessable (*check for plagarism and accuracy), but process (2), which leads to chargino production in cascade decays, will be dominant if the gluino mass is within reach. The ISR jet in process (1) is not physically required; it simply provides something for the gluinos to recoil off of so the MET signal isn't lost (*My interpretation. Run it by someone who knows stuff).

@@

\subsection{Stopped Particles}
            (*cite CMS PAS EXO-17-004)
            (*mention how stopped particles complement HSCP search?)
            (*Allen talks about trigger efficiency stuff. Is that really important?)
            With long enough lifetimes and sufficient mass, LLPs could actually come to rest in the detector before decaying. The eventual decay would then lead to energy deposits in the calorimeters or hits in the muon system that could be out of time with proton bunch crossings. A recent CMS analysis looked for such a signal in dimuon final states (*necessarily dimuon?) using the full 2015 and 2016 datasets.

            This analysis is interpreted in the context of two potential LLPs: gluinos and multiply charged massive particles (mchamps). Long-lived gluinos can appear in models such as Split SUSY, where the squarks and sleptons are quite heavy, but the gluino is relatively light. After being pair produced, gluinos could form R-hadrons and eventually decay to quark-antiquark pairs along with NLSP neutralinos, which then decay to LSP neutralinos along with muon-antimuon pairs. The gluino decay is supressed by the large squark mass, which leads to its appreciable lifetime. Mchamps will not be discussed further because their existence is not directly motivated by naturalness (*is this true?).

@@

            HSCPs will often be misidentified or unobserved by standard reconstruction algorithms and analysis criteria, which tend to assume beta~1 and |q|=+-e (*dE/dx plot is nice around here). Inverting these assumptions provides two handles for a dedicated HSCP search in the context of three benchmark models: (1) HSCPs that form R-hadrons that either change the sign of their electric charge or become neutral before reaching the MS, (2) lepton-like HSCPs (quasi-stable sleptons) that are produced either through decays of squarks or gluinos or directly if squarks and gluinos are kinematically inaccessible, and (3) long-lived leptonlike fermions with aritrary electric charge. This discussion will focus on the first two models, as they bear more directly on naturalness and SUSY.

@@

            The authors estimate the signal-region background using and ABCD method in which events are binned in a 2D plane with uncorrelated axes and the ratio of events in two control regions is used to extrapolate into the signal region from a third control region (*Do I need to explain this? If I do, I should be more clear). Systematic uncertainties are introduced in background estimation, signal acceptance, and integrated luminosity.

>>>>

    (*kinda funny wording throughout. Maybe it would be better to trade some words for math?)
    All SM particles interact with virtual particles through loop interactions. These interactions lead to corrections to SM parameters such as particle masses. It is common to write a given parameter as a linear combination of its 'bare' (uncorrected) value and its corrections. As the only fundamental scalar in the SM, the Higgs is uniquely affected by these corrections. The next section explores the concept of naturalness, a major motivating principle for beyond the standard model (BSM) physics that is directly tied to the relationship between the Higgs bare mass and the terms that correct it.

\section{Naturalness and the Higgs mass}
    The naturalness criterion takes several forms. Here are a few (*cite these): (1) all parameters in a model should be of order 1, (2) Loop-level contributions should not be larger than tree-level contributions, (3) any small parameter should be protected by a custodial symmetry, and (4) fine-tuning is bad (*probably a better way to say 4). Although these statements are all motivated by the samefundamental notion, I find (3) and (4) to be most compelling.

@@

    (*talk about anthropic principle or other arguments against?)
    (*I got tired of reading this section and haven't edited this paragraph at all yet)
    There are a few proposed solutions to the SM naturalness problem. Supersymmetry is the main one, and I'll spend the great majority of my time on it. Extra dimensions and composite Higgs models are the other two that pop up fairly often, but won't spend much time on them outside the next two sentences. Extra dimensions solves the hierarchy problem either by lowering the cutoff scale (as in large extra dimensions (ADD)) or introducing a warping factor that reduces the level of unnaturalness (warped extra dimensions (RS)). Composite Higgs models propose that the Higgs stops behaving as a fundamental scalar at some energy level, which removes the quadratic dependence. (*Hidden valley, dark sector, little Higgs?)

\section{Supersymmetry}
    Supersymmetry (SUSY) introduces a new symmetry in which every SM field fits into a larger multiplet with an inherent symmetry between bosons and fermions. In it's simplest form, SUSY results in one new boson for every SM fermion and one new fermion for every SM boson (*and an extra Higgs field - can I gloss over that?). The increase in particle multiplicity calls for a new naming convention: the spin-0 superpartners to the SM fermions are referred to as sfermions (e.g. stop or smuon), and the spin-1/2 superpartners to the SM bosons are given the name of their SM counterpart with "ino" tacked on to the end (e.g. Higgsino or wino). Together the superpartners to the SM particles are known as sparticles.

    When calculating contributions from loop diagrams, one finds that fermion loops differ in sign from boson loops, which means that in SUSY every bosonic correction to the Higgs mass is cancelled by a fermionic correction and vice versa. In fact, exact supersymmetry predicts that the bare Higgs mass is uncorrected to all orders in perturbation theory. (*show example diagram with contribution?)

    But the lack of sparticles in collider searches and everyday life tells us that SUSY must be broken. If SUSY were exact, every superpartner would have the exact same mass as its SM counterpart, which means all sparticles would have been found at the same energy scales as SM particles. The mechanism and scale of SUSY breaking is an open question with many competing answers that achieve different theoretical goals and lead to different TeV-scale phenemena. A handful of models with particular approaches to SUSY breaking that are particularly prominant or particularly relevant to naturalness or long-lived particles are gauge-mediated susy breaking (GMSB), anomaly-mediated SUSY breaking (AMSB), Split SUSY, and R-parity violating SUSY (RPV). The details of these models will be explored as needed throughout this paper.

\section{Long-lived Particles}
    The SUSY variants most relevant to this discussion are those that are motivated by naturalness and (*and/or?) produce a signal that could be missed by common SUSY search techniques. Specifically, we are interested in SUSY models that predict long-lived particles (LLPs). That is, particles whose lifetimes are such that they decay at least a detectable distance from the nominal interaction point. This category includes everything from particles that decay at (*give ballparky number (~100um?)) to particles that propagate through the entire detector.

    In the SM and in BSM theories, LLPs result from one of the following conditions: (1) a symmetry that guarantees stability, (2) weak couplings to decay products, (3) highly-virtual intermediate states, or (4) limited phase space in which to decay. In the SM, charge and baryon number conservation guarantee the absolute stability of the electron and proton as the the lightest particles carrying their respective conserved quantities. Neutrons and muons, on the other hand, are long-lived beause the proton and electron masses leave them little phase space in which to decay (*also mention that they're both weak decays?). (*Mention neutrino, pion, kaon, some other bound state I might be forgetting?)

    In SUSY models, LLPs arise for the same basic reasons, and different SUSY models meet different LLP criteria in different ways. The most common (*based on my assumptions) SUSY LLP occurs in models with conserved R-partity, which is a proposed symmetry that protects many models from proton decay. In models with R-parity, the SM particles are assigned R-parity +1 and their superpartners are assigned R-parity -1. Conserving the product of R-parity at each vertex has two immediate phenomenalogical consequences: (1) sparticles can only be produced in pairs and (2) the lightest supersymmetric particle (LSP) is absolutely stable. A neutral LSP could show up in collider searches as significant missing energy, making MET the conventional SUSY signature in many searches.

    On the other hand, models with R-parity violation (RPV) produce long-lived, but not perfectly stable LSPs when RPV terms have small coupling constants. Luckily, RPV models can still avoid proton decay by allowing either terms that violate baryon number or terms that violate lepton number, but not both. A similar situation arises in GMSB models where the gravitino is the LSP. The coupling between the next-to-LSP (NLSP) and gravitino is inversely proportional to the scale of SUSY breaking, which supresses the NSLP decay rate and can lead to LLPs. (*is now a good time to explain GMSB?)(*cite dimopoulos low energy)

    LLPs from highly-virtual intermediate particles can occur in models such as split SUSY, where the scalar sparticles are significantly more massive than the gauginos. In these models, the gluino becomes long-lived when its decay to a quark and neutralino is mediated by a highly-virtual squark. (*cite kilian) (*mention that split throws out naturalness motivation?)

    The degenerate mass spectra of other SUSY models can also give rise to LLPs. In some AMSB models (*explain AMSB?), for example, the zino and wino are the LSP and NLSP, respectively, and are nearly degenerate in mass. The wino then becomes long-lived when its decay is suppressed by the lack of available phase space. (*cite R\&S out of this world)
    (*merge last 2 paragraphs under 'mass spectra lead to LLP' concept?)
    (*never said anything about hidden valley or dark sector stuff. Never said the word R-hadron. Never said stealth or displaced SUSY. Should I?)

@@

\subsection{Tracker}
    (*cite CMS TDR, jinst 12 c02033)
    (*maybe I should trade some number listing for talking about how silicon detectors work)
    (*what about readout, cooling, etc?)
    In the region closest to the interaction point, CMS uses a fast, high-granularity silicon tracker to reconstruct tracks and pinpoint primary and secondary vertices. The tracker is subdivided into two regions: at r < 20 cm, the high particle flux (*specify) requires small pixels to keep occupancy near 0.1\%. Moving farther from the IP decreases particle flux, which allows the use of larger silicon strips.

@@

    Just outside the pixel tracker, the silicon strip tracker (SST) occupies the 20 < r < 116 cm and |z| < 282 cm region of CMS. Composed of 9.3 million silicon strips that vary from \_ to \_ in size (*look up) depending on particle flux in the region, the SST covers an active area of roughly 196 m\^2. At each trigger, the strip output voltage is transferred via optical link to be read out by off-detector front-end electronics (*right jargon?).

\subsection{Electromagnetic Calorimeter}
    (*cite CMS TDR)
    (*talk about readout?)
    (*More in the way of concept/how stuff works?)
    After traversing the inner tracker, a particle would next encounter the electromagnetic calorimeter (ECAL). As a homogeneous scintillation calorimeter, ECAL uses 61200 lead tunstate crystals in the barrel and 7324 in each endcap to reconstruct the energy deposited during electromagnetic showers. Lead tungstate crystals allow for a fast (80\% of light within 25 ns), compact (radiation length = 0.89 cm), fine-grained (Moliere length = 2.2 cm), and radiation hard (up to 10 Mrad(*learn context for this number)) calorimeter. The main drawback is the relatively low light yield (30 photon / MeV), which necesitates photodetectors with intrinsic gain that work in magnetic fields (*say B-field part better).

    The barrel section (EB) extends radially from 129 to 177 (*look up) cm and covers a pseudorapidity range of 0 to 1.479. The crystals are tapered to approximately project back to the IP but not so perfectly that likely particle trajectories align with cracks. Each crystal is approximately one Moliere radius wide and 25.8 radiation lengths deep.

@@

    Each endcap spans a pseudorapidity range of 1.3 to 3.0 with 14 towers in eta and 5 to 10 degree phi segmentation. Also, a steel and quartz fiber forward calorimeter (HF) sits 11.2 m from the interaction pipe in a pseudorapidty range of 3.0 to 5.0. In HF, particles produce Cherenkov light when traversing the quartz fibers that run parallel to the beamline. (*say why HF is useful or mention crazy flux?)

@@

\subsection{Trigger}
    (*cite CMS TDR, jinst 12 c03021)
    (*Is entire detector read out at L1?)
    (*Mention fpgas in L1 or more details about HLT?)
    (*Something about particle flow?)
    (*Sloppy, but I'm getting tired)
    The trigger reduces the data writing rate from the 40 MHz collision rate to less than 1 kHz so that events can be written to tape. The rate reduction happens in two stages: Level-1 (L1) and High-Level Trigger (HLT). L1 uses input from ECAL, HCAL, and MS that represents physics objects (e/gamma, tau, jets, MET, and mu) and uses custom electronics to reduce the rate to ~100 kHz in 3.8 us. The HLT then uses a dedicated processor farm to further reduce the rate to the desired < 1 kHz.

\section{LHC Searches for Long-lived Particles}
\subsection{Overview}
    LLPs leave distinct signatures that differentiate LLP signals from SM backgrounds but also add complexity to LLP analyses. CMS and ATLAS were generally designed to detect and reconstruct stable particles from prompt decays, so LLP analyses often need to stray from the traditional analysis path in terms of object reconstruction (*really reconstruction? Maybe object definition?), background estimation, and which data tier to use (*can I say this more generally?). This paper will focus on four complementary LLP signatures: displaced vertices, disappearing tracks, slow-moving particles, and particles that stop completely before decaying.

    Displaced vertices and disappearing tracks both look for particles that are unstable on detector timescales (*quantify?). Displaced vertices searches are sensitive to LLPs that decay to visible particles in the inner volume (*specify) of the detector and leave tracks that can be traced back to a vertex that is sufficiently displaced from the nominal interaction point (*wordy). Disappearing track searches are sensitive to LLPs that decay to particles that are either neutral or too soft to be reconstructed. Together, these two types of analyses cover (*something something) parameter space of LLPs that decay on detector timescales.

    Slow and stopped particle searches are both sensitive to LLPs that are detector-stable but far more massive than any stable SM particle. Searches for low-velocity particles use ionization energy loss per unit distance (dE/dX) and time-of-flight (TOF) measurements to look for particles that are moving more slowly than the expected SM particles (*redundant), which normally (*quantify) have beta > 0.9. Stopped particle searches take this idea one step further and look for particles that actually come to rest inside the detector before decaying, a phenomenon that could be possible with long lifetimes and small initial kinetic energy. These analyses look for calorimeter energy deposits or hits in the muon system that are out of time with proton bunch crossings that could be due to the decay of stopped particles.

@@

\subsection{Displaced Vertices}
    (*cite CONF-2017-026)
    LLPs that decay before reaching the inner tracker but still live longer than the vast majority of SM particles can be identified with a displaced vertex (DV). That is, a reconstructed vertex far from the nominal interaction point. ATLAS performs a search for such particles by looking for at least one DV along with large MET and large track multiplicity. The search is sensitive to LLPs with lifetimes in the 20 ps to 10 ns range.

@@

    Jets and MET are reconstructed as in standard analyses, but special care must be taken when looking for displaced tracks. The standard ATLAS tracking algorithm constrains the transverse and longitudinal impact parameters of potential tracks, so the authors employ an additional algorithm, referred to as large-radius tracking (LRT). LRT is run after the standard tracking algorithm and uses only hits that were not used in standard track reconstruction. By relaxing d0, z0, and number of hits requirements, LRT is able to reconstruct tracks that point back to DVs.

@@

    Both ATLAS and CMS looked for disappearing tracks during Run I, but the tracker improvements installed in ATLAS before 2015 and CMS before 2017 should increase sensitivity to decays at shorter radii (*I haven't mentioned previous iterations when talking about other analyses). The ALTAS search presented here is sensitive to LLP lifetimes of approximated 10 ps to 10 ns (*make sure I'm talking about lifetime sensitivity in the same way across analyses).

    The authors consider two signal processes: (1) electroweak gaugino production resulting in a disappearing track, an ISR jet, and MET and (2) gluino pair production resulting in a disappearing track, four jets, and MET. Process (1) could be the only gaugino production mode at the LHC if the gluino mass is kinematically unaccessable, but process (2), which leads to chargino production in cascade decays, will be dominant if the gluino mass is within reach. The ISR jet in process (1) is not physically required; it simply provides something for the gluinos to recoil off of so the MET signal isn't lost (*My interpretation. Run it by someone who knows stuff).

@@

\subsection{Stopped Particles}
    (*cite CMS PAS EXO-17-004)
    (*mention how stopped particles complement HSCP search?)
    (*too redundant with overview?)
    With long enough lifetimes and sufficient mass, LLPs could come to rest in the detector before decaying. The eventual decay would then lead to energy deposits in the calorimeters or hits in the muon system that could be out of time with proton bunch crossings. A recent CMS analysis looked for such a signal in dimuon final states (*necessarily dimuon?) using the full 2015 and 2016 datasets.

    This analysis is interpreted in the context of two potential LLPs: gluinos and multiply charged massive particles (mchamps). Long-lived gluinos can appear in models such as Split SUSY, where the squarks and sleptons are quite heavy, but the gluino is relatively light. After being pair produced, gluinos could form R-hadrons and eventually decay to quark-antiquark pairs along with NLSP neutralinos, which then decay to LSP neutralinos along with muon-antimuon pairs. The gluino decay is supressed by the large squark mass, which leads to its appreciable lifetime.

@@

    HSCPs will often be misidentified or unobserved by standard reconstruction algorithms and analysis criteria, which tend to assume beta~1 and \|q\|=+-e (*dE/dx plot is nice around here). Inverting these assumptions provides two handles for a dedicated HSCP search in the context of three benchmark models: (1) HSCPs that form R-hadrons that either change the sign of their electric charge or become neutral before reaching the MS, (2) lepton-like HSCPs (quasi-stable sleptons) that are produced either through decays of squarks or gluinos or directly if squarks and gluinos are kinematically inaccessible, and (3) long-lived leptonlike fermions with aritrary electric charge. This discussion will focus on the first two models, as they bear more directly on naturalness and SUSY.

@@

    The authors estimate the signal-region background using an ABCD method in which events are binned in a 2D plane with uncorrelated axes and the ratio of events in two control regions is used to extrapolate into the signal region from a third control region (*Do I need to explain this? If I do, I should be more clear). Systematic uncertainties are introduced in background estimation, signal acceptance, and integrated luminosity.
